Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-17 paper.tex
Tone down banter

We believe that there is also a broader take-away for an audience of
French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because our government-approved programming language does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete data structure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. fully exploit parametricity for
reasoning.

@@

Unfortunately, \coqe!shiftrows_! and \coqe!Rows.ror! would entail some
very fiddly and computationally costly bit twiddling. These amount to
performing bit-level permutations within a machine word. This is a
frequent pain point when bitslicing the diffusion layer of ciphers: it
is also true in the case of \Gift{} and \AES{}, for instance. In the
next section, we present this one weird trick to get permutations to
nearly vanish.

>>>>

We believe that there is also a broader take-away for an audience of
French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because OCaml, our favorite  programming language, does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete data structure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. fully exploit parametricity for
reasoning.

@@

Unfortunately, \coqe!shiftrows_! and \coqe!Rows.ror! would entail some
very fiddly and computationally costly bit twiddling. These amount to
performing bit-level permutations within a machine word. This is a
frequent pain point when bitslicing the diffusion layer of ciphers: it
is also true in the case of \Gift{} and \AES{}, for instance. In the
next section, we present a clever trick to get permutations to
nearly vanish.
